---
title: "Census Data"
author: "Charmayne Patterson"
date: "7/23/2019"
output:
  html_document: default
  pdf_document: default
---

## Introduction
In order to explore the potential relationships between several variables related to the status of African Americans in Philadeplhia in 1847, I must start off by reading the census data.    

```{r, warning=FALSE, message=FALSE}
library(tidyverse) #The swiss knife!
library(RCurl) #To read the csv file from GitHub directly
```


Now that the packages  are loaded,  I can get the data into R. 

```{r}
census <- read.csv("census3.csv")
```

## Exploratory Data Analysis

Now that we have ingested the data, let us start exploring it. The first thing to do is to have a quick look at our data. For that, we can use the `str()` function or the more human readable (and intuitive), `glimpse()` function.

```{r}
glimpse(census)
```

Now I can begin exploring the data. The first thing to do is to have a quick look at it, noticing that there are 2851 rows in our dataset with 11 variables. It added the `X1` variable on its own while importing the data. Not only can I see my variables, but I also now know the type of variables, for example, `Male Occupation` is a characteer variable. Should it be a factor?based on the numerous observations and 11 variables, it is obvious that the data must be cleaned in order to make it more manageable and readable. 



While carrying out any analysis, it is always best to start off with a question. It is also important to know the metadata, or the data that describes the data. For example, I see the variable `Can.read` has values ranging from 0 to 5 (we will of course check this), but I have no idea what these values mean. I would have assumed that this might be a binary value with 0 indicating the person cannot read and 1 indicating that the person can read. Or this could also mean the total number of people in a household who can read. That sounds more plausible to me. Either way, I will look at the metadata or a data dictionary, if one exists. I can also observe the different values available in the `Can.read` and `Can.write` variables.

```{r}
table(census$Can.read)
table(census$Can.write)
```
Looking at this, I am now more inclined to think that these two variables show me the number of people who can read or write in a family. So with these assumptions, what are the questions that come to my mind? Personally I would really like to understand the impact of proffessions, gender and other variables available, on the  level of education. For the 'level of education' I have two variables available, viz. `Can.read` and `Can.write`. I also know the size of a family. So perhaps it would be best to divide these two variables with the size of the family and look at this as a percentage to say what percent of a family is well versed with reading or writing. Prior to that I also want to check for discrepancies in the dataset. Given the family size, the sum of the Males and Females should be equal to the family size, considering this era only provided binary options for gender. 


```{r}
#Create a new variable called GenderSum and then look at the difference
census <- census %>% mutate(
  GenderSum = Males + Females,
  FamDiff = `Family.Size` - GenderSum
)

which(census$FamDiff > 0)
```

There are about 21 values that don't quite match up.Row 71 provides an example. While it says that the family size is 5, it has only 1 male and 2 females. This doesn't quite add up. So maybe removing these values and the two additional columns we created, as well as the `X1` column.



```{r}
census <- census %>% filter(FamDiff == 0) %>% select( -GenderSum, -FamDiff)
```

Now I can mutate a new variable baased on the percentages discussed above.

```{r}
census <- census %>% mutate(
  Read = round((Can.read/`Family.Size`)* 100, 2),
  Write = round((Can.write/`Family.Size`)* 100, 2)
)
```


I could also perhaps combine these two newly created variables and have one cummulative variable which is just the mean of these two, to signify the overall education level. After creating this new variable I won't need the original variables.


```{r}
census <- census %>% mutate(
  EduLevel = round((Read + Write)/2, 2) 
) 
```

I must check and see that are `EduLevel` is not greater that 100%.

```{r}
which(census$EduLevel > 100)
```


Looking at row 144, I see that the family size is 3, but the number of people who can read or write is 6 and 4 respectively. This alerts me to the fact that something is wrong here. I will need to filter for only those values that are less than or equal to 100.

```{r}
census <- census %>% filter(EduLevel <= 100) %>% select(-(Can.read:Write))
```

Ok, so far so good. Now I must address the task of making some sense of the male and female occupations. There are too many occupations and so they need to be reduced. The remainder of this analysis focuses on education level, literacy, and occupations.  The data frame (census_glm.csv) below includes a variable labeled literate.  If a family has above a 50% education level, then the family is literate.  A (1) in the literate column specifies that a family is literate and a (0) means that the family is not literate.    


```{r}
census_glm<- read.csv("census_glm.csv")
```

Additional variables are included in the data file.  These variables are removed using the code that follows.

```{r}
census_glm<-select(census_glm, -c(X.1, X))
```

Using plots and graphs, we explore the data with the nine questions in mind.These questions can help me to think about the data.

1. What was the most common profession for males? 
2. Waht was the most common profession for females? 
3. How do these professions measure up when viewed from the education level? 
4. Which is the least educated profession? 
5. Which is the most educated profession? 
6. Is there a relationship between family size and Education Level? 
7. Is there any diﬀerence between genders as far as education level is concerned? 
8. Does being born a slave have any bearing on the education level? 




```{r}
census_glm %>% group_by(Male.Occupation) %>% count() %>% arrange(desc(n))
census_glm %>% group_by(Female.Occupation) %>% count() %>% arrange(desc(n))
```

The occupation with the largest number of males is porter. Since there are 1,426 different occupations, a filter is created to include only occupations that have at least 20 male family members.  When viewing the occupations for females, it seems that more females seem to hold a position than males.  This observation was fascinating to me as the male in modern times are viewed as the family provider.  The most popular female position held is a washer.  There are 392 women washers in the data set. Since there are 796 different occupations, a filter is created to include only occupations that have at least 20 female family members.The results are printed to the consoloe.





```{r}
male.occ_count <- census_glm %>% group_by(Male.Occupation) %>% count() %>% filter(n > 19) %>% arrange(desc(n))
female.occ_count <- census_glm %>% group_by(Female.Occupation) %>% count() %>% filter(n > 19) %>% arrange(desc(n))
male.occ_count
female.occ_count

```


This filter seems to be a good one. Now take a deeper dive into the data to determine education level by occupation. A comparison among occupations answers question 1. The results will be displayed as a bar graph.

```{r}
ggplot(census_glm %>% filter(Male.Occupation %in% unique(male.occ_count$Male.Occupation)),
       aes(x = Literate, fill = Male.Occupation)) + geom_bar(position = 'dodge')

ggplot(census_glm %>% filter(Female.Occupation %in% unique(female.occ_count$Female.Occupation)),
       aes(x = Literate, fill = Female.Occupation)) + geom_bar(position = 'dodge')
```


Of the 14 chosen male occupations, families that have male porters have the highest illiterate level.  If a family has a male waiter, then the family is more likely to be literate.  Females that hold a position washing and ironing have the least educated families.  However, families that have female washers have higher family literacy levels.


Attention is shifting to determining a relationship between the variables.   We observe the relationship between family size and education level.  A color scheme is used to distinguish between the literate and non-literate group easily.


```{r}
ggplot(census_glm, aes(x = EduLevel, y = Family.Size, color=Literate)) +

  geom_point() + geom_jitter()
```


An interesting relationship between the variables is recognized.  The relationship seems to be parabolic.


```{r}
ggplot(census_glm, aes(x = EduLevel, y = Females, color=Literate)) +

  geom_point() + geom_jitter()

ggplot(census_glm, aes(x = EduLevel, y = Males, color=Literate)) +

  geom_point() + geom_jitter()
```








```{r}
ggplot(census_glm, aes(x = Born.slaves, y = EduLevel, color=Literate)) +

  geom_point() + geom_jitter()
```


Consider question number 8, we draw a table to capture information to answer the question and to confirm the results from the graph above.

```{r}
census_glm %>% group_by(Literate, Born.slaves) %>% count() 

```


Out of 2242 study participants, 1710 were born slaves and are in a household that is deemed to be illiterate while 514 of the participants that were taken as slaves have literate families.  Occupation is discussed next.



